---
title: "Medical Charges: Data Analysis"
author: "Group - 7"
date: "25/11/2021"
output:
  html_document: default
  pdf_document: default
  word_document: default
---
tinytex::install_tinytex
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

We begin with importing the ggplot2 and ggpubr packages required for data visualization. We also read the data from the .csv file containing insurance data.

```{r}
library(ggplot2)
library(ggpubr)
library(corrplot)
insurance = read.csv("insurance.csv")
```
Let us now see the structure and summary of the given dataset.

```{r}
str(insurance)
```
```{r}
summary(insurance)
```

$$\text{Checking for Missing Values in the Data}$$

```{r}
f=function(x){any(is.na(x))}
check.na=apply(insurance,2,f);
check.na
```

As we can see in the above output, no column in the Insurance dataset contains empty value. Hence, there are no missing values in the dataset.


$$\text{Exploratory Data Analysis}$$

Clearly, charges is the output variable in this dataset. It gives us the insurance amount for a resident (a row of input variables in the dataset). Let us examine the distribution of charges using histogram. 

```{r}
insurance %>%
ggplot(aes(charges)) +
geom_histogram(color = "blue", fill = "purple",alpha = .5, bins = 25) +
scale_x_continuous(breaks = seq(0,66000,10000)) +
theme(axis.text.x = element_text(size = 14)) +
labs(title="Charges", x="")
```

The histogram plot shows that the distribution of charges is right-skewed. Let us confirm it by calculating the skewness of charges. 

```{r}
library(moments)
charges = insurance$charges
skewness(charges)
```
As expected, the skewness measure is positive. It indicates that the charges are positively (right) skewed.

Let us begin by visualizing the relation between the charges incurred by a resident and their region.  

```{r}
ggplot(insurance) +
  aes(x = region, y = charges, fill = region) +
  geom_boxplot(shape = "circle") +
  scale_fill_brewer(palette = "PRGn", direction = 1) +
  labs(
    x = "Region",
    y = "Charges",
    title = "Boxplot of Medical Charges for different regions"
  ) +
  theme_minimal()
```


These boxplots indicate that the average medical cost is similar for all the regions. That is, a resident's region does not have much impact on the medical cost incurred. 

Let us now examine if smoking affects the charges for a person. 
```{r}
ggplot(insurance) +
  aes(x = smoker, y = charges, fill = smoker) +
  geom_boxplot(shape = "circle") +
  scale_fill_hue(direction = 1) +
  labs(
    x = "Smoking Status",
    y = "Charges",
    title = "Boxplot of medical charges as per Smoking Status"
  ) +
  theme_minimal()
```

Clearly, the residents who smoke have higher charges than those who do not. This observation is further supported by separating boxplots for male and female residents.

```{r}
ggplot(insurance) +
  aes(x = smoker, y = charges, colour = sex) +
  geom_boxplot(shape = "circle", fill = "#112446") +
  scale_color_hue(direction = 1) +
  labs(
    title = "Boxplot of Smoking Status & Sex of the person."
  ) +
  theme_minimal()
```

Even in the population that smokes, the male residents incur higher charges than femalle residents.

Now, we will check if the number of children a person has affects their medical charges.

```{r}
# encoding number of childern as categorical variable
insurance$children =  as.factor(insurance$children)
ggplot(insurance) +
  aes(x = children, y = charges, fill = children) +
  geom_boxplot(shape = "circle") +
  scale_fill_brewer(palette = "Paired", direction = 1) +
  labs(
    x = "Children",
    y = "Medical Charges",
    title = "Boxplot of number of children vs medical charges"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

We see that the mean charges for a person is the least when they have five children.

$\textbf{Impact of BMI on insurance charges: }$

```{r}
obesity = ifelse(insurance$bmi >= 30,1,0)
overweight = ifelse(insurance$bmi >= 25,1,0)
insurance_updated = cbind(insurance, as.factor(obesity), as.factor(overweight))

ggplot(insurance_updated) +
  aes(
    x = `as.factor(overweight)`,
    y = charges,
    fill = `as.factor(obesity)`
  ) +
  geom_boxplot(shape = "circle") +
  scale_fill_manual(
    values = c(`0` = "#8E0152",
    `1` = "#276419")
  ) +
  labs(
    x = "Overweight",
    y = "Medical Charges",
    title = "Effect of Obesity & Overweight on Medical Charges"
  ) +
  theme_minimal()
```

$\textbf{Relation between charges and age}$ : 

Let us first classify the population into different age groups: less than $30$ years, $31$ to $40$ years, $41$ to $50$ years, $51$ to $60$ years and $61$ to $70$ years 

```{r}
summary(insurance$age)
# we will bucket people as per age group and see how the medical charges trend in those groups.
age_group = vector("character")
for(i in 1:nrow(insurance)){
  if(insurance$age[i] <= 30){
    age_group[i] = "18-30"
  }else if (insurance$age[i] <= 40){
    age_group[i] = "31-40"
  }else if (insurance$age[i] <= 50){
    age_group[i] = "41-50"
  } else if(insurance$age[i] <= 60){
    age_group[i] = "51-60"
  }else if (insurance$age[i] <= 70){
    age_group[i] = "61-70"
  } 
}
table(age_group)
age_group = as.factor(age_group)
insurance_updated = cbind(insurance_updated, age_group)

ggplot(insurance_updated) +
  aes(x = age_group, y = charges, fill = age_group) +
  geom_boxplot(shape = "circle") +
  scale_fill_hue(direction = 1) +
  labs(
    x = "Age Group",
    y = "Medical Charges",
    title = "Effect of age on medical charges"
  ) +
  theme_minimal() +
  theme(
    axis.title.y = element_text(face = "bold"),
    axis.title.x = element_text(face = "bold")
  )
```

The boxplots indicate that the average medical charges are higher for the population in higher age groups. 

```{r}
ggplot(insurance_updated, aes(x=age, y=charges)) + geom_point()
```

$$\text{Correlation Between Different Columns in the Dataset}$$

```{r}
# make correlation table
str(insurance)
insurance$children = as.numeric(insurance$children)
# encoding yes as 1 and no as 0
# female as 0 and male as 1
insurance$smoker= ifelse(insurance$smoker == "yes", 1,0)
insurance$sex = ifelse(insurance$sex == "male", 1 , 0)
# encoding southwest as 1, southeast as 2, northwest as 3, northeast as 4
region_encode = vector("numeric")
for(i in 1:nrow(insurance)){
  if(insurance$region[i] == "southwest"){
    region_encode[i] = 1
  }
  else if (insurance$region[i] == "southeast"){
    region_encode[i] = 2
  }
  else if(insurance$region[i] == "northwest"){
    region_encode[i] = 3
  }else if (insurance$region[i] == "northeast"){
    region_encode[i] = 4
  }
}
insurance_updated = cbind(insurance, region_encode)
head(insurance_updated)
colnames(insurance_updated)
cor(insurance_updated[,-6])
corrplot(cor(insurance_updated[,-6]), method = "color")
```

The correlation plot indicates that charges are mildly correlated to age and BMI of a person, and strongly correlated to whether they smoke or not. We classify the population into obese and overweight to see how the charges fare for both the categories.

$$\text{Building Models to Predict the Charges}$$


We can now build models to predict the charges for a resident. 

```{r}
# Now we build the model.
# Split the data into train & test.
set.seed(123)
ID = 1:nrow(insurance)
insurance = cbind(ID, insurance)
train = sample(ID,1000)
test = ID[-train]
train = insurance[train,-1]
test = insurance[test,-1]
lr = lm(charges~., data = train)
summary(lr)
```

We remove the following variables from our original model as they are insignificant: $\textbf{sex}$ and $\textbf{region}$.
```{r}
lr2 = lm(charges~ age + bmi + children + smoker, data = train )
summary(lr2)
```

The adjusted R square decreases by a very small margin, but we are able to simplify the model by removing two variables.

```{r}
c = lr2$coefficients
p = vector("numeric")
for (i in 1:nrow(test)){
  p[i] = c[1] + test$age[i]*c[2]+test$bmi[i]*c[3]+ test$children[i]*c[4]+test$smoker[i]*c[5]
}
residual = (test$charges-p)
plot(residual)
#The RMSE is:
sqrt(mean(residual^2))

```
PREDICTION $1$ : $19$-year old female, smoker, overweight and with no children, resident of the Southwest region of the USA.
```{r}
# Prediction 1
predict(lr2, data.frame(age = 19, bmi = 27.9 , children = 0, smoker = 1))
```


PREDICTION $2$ : $55$-year-old man, smoker, non-obese and without children, resident of the Northeast region of the USA.
```{r}
# Prediction 2
predict(lr2, data.frame(age = 55,bmi = 25, children = 0, smoker = 1))
```



PREDICTION $3$ : $70$-year-old woman, non-smoker, obese and with two children, resident of the southeastern USA.

```{r}
# Prediction 3
predict(lr2, data.frame(age = 70, bmi = 35, children = 2, smoker = 0))
```
PREDICTION $4$ : $22$x-year-old woman, smoker, non-obese and with 4 children, resident of the northwestern region of the USA.

```{r}
# Prediction 4
predict(lr2, data.frame(age = 22,bmi = 23, children = 4,smoker = 1))
```

$$\text{Conclusion}$$

We carried out data analysis using  library and also linear regression on charges and other input variables. We found that: 

1. The variable $\textbf{smoker}$ affects charges the most. That is, a person who smokes is likely to incur higher charges than those who do not.

2. The BMI of a person mildly affects the charges. If a person is in obese category, they are expected to incur higher charges than other overweight and underweight residents. 

3. On an average, people with 5 children are charged less than others.

4. Region and sex of a resident do not affect the charges significantly. Hence, these input variables are not of much interest while studying the insurance data.

We built two models: by filtering on the basis of significance. 

1. Regressing charges on all the input variables
2. Regressing charges on all the input variables $\textbf{except sex and region}$ .

In the second model, the fit is not impacted adversely as the adjusted $R^2$ value decreases slightly from $0.7469$ to $0.7468$.

